---
title: "Group Project 2"
author: "Paul Lichtenfeld"
date: "2024-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("knitr", "kableExtra", "magrittr", "readr", "geosphere", "plyr", "dplyr", "kable", "tidyverse", "tibble", "ggplot2","reshape2","stringr","leaflet")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
#if (length(install_me)) install.packages(install_me)

library(knitr)
library(geosphere)
library(plyr)
library(dplyr)
library(kableExtra)
library(tibble)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(stringr)
library(leaflet)
devtools::install_github("haozhu233/kableExtra") 
```

## Objective 1 
The objective was to create an interactive map for both the global death dataset and the global confirmation dataset. The first thing I did was read in the two data frames. Next I modified the death data set. First I set a variable to be the column name of the final column in the data set. I then created a new variable which held the data set grouped by the country, averaged out the latitude and longitude for countries with multiple entries, and summed the total number of deaths for countries with multiple entries. I then did the same for the confirmed data set. I then made two data frames which held the labels, latitude, longitude, and the total number of confirmations/deaths for the visualization. I then wrote the color palettes for the two map layers. I then created the visualization, passed it map tiles and set two different toggleable groups, one for the confirmed data set and one for the death data set. 

```{r Obj1}
#two necessary data frames
deathGlobal<- read.csv("time_series_covid19_deaths_global (1).csv", header = TRUE)
confirmedGlobal <- read.csv("time_series_covid19_confirmed_global (2).csv", header = TRUE)

#modifies the death data set to what I need 
finColNameDeath <- (colnames(deathGlobal[ncol(deathGlobal)]))
newDeath <- deathGlobal %>% group_by(Country.Region) %>% summarize(averageLat=mean(Lat, na.rm=TRUE), averageLong=(mean(Long,na.rm = TRUE)), totalDeaths=sum(.data[[finColNameDeath]]))


#modifies the confirmed dataset to be what I need it to be 
finColNameConfirm <- colnames(confirmedGlobal[ncol(confirmedGlobal)])
newConfirm <- confirmedGlobal %>% group_by(Country.Region) %>% summarize(averageLat=mean(Lat,na.rm = TRUE),averageLong=mean(Long,na.rm = TRUE),totalConfirms=sum(.data[[finColNameConfirm]]))



#Creates a data frame for the visualization 
confirmLocations <- data.frame(
  label = (deframe(newConfirm[,1])),
  latitude = deframe(newConfirm[,2]),
  longitude = deframe(newConfirm[,3]),
  totalConfirms = deframe(newConfirm[,4])
  
)
#creates a data frame for the death part of the visualization 
deathLocations <- data.frame(
  label = deframe(newDeath[,1]),
  latitude = deframe(newDeath[,2]),
  longitude = deframe(newDeath[,3]),
  totalDeaths = deframe(newDeath[,4])
)

#color pallette function call 
pal <- colorFactor(c("navy","purple","red"), domain= newConfirm$totalConfirms)
secPal <- colorFactor(c("navy","purple","red"), domain= newDeath$totalDeaths)

#creates interactive visualization 
leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>%
  #first circle group for the confirmed data set 
  addCircles(
    data = confirmLocations,
    lat = ~latitude,
    lng = ~ longitude,
    popup = ~as.character(totalConfirms),
    label = ~label,
    radius = 190000,
    stroke = FALSE, fillOpacity = 0.5, 
    color = ~pal(totalConfirms),
    group = "Confirmations"
  )%>%
  addCircles(
    #second circle group for the death data set 
    data = deathLocations,
    lat= ~latitude,
    lng = ~longitude,
    popup = ~as.character(totalDeaths),
    label = ~label, 
    radius = 190000,
    stroke = FALSE, fillOpacity = 0.5, 
    color = ~secPal(totalDeaths),
    group = "Deaths"
  )%>%
  #controls the two possible layers for the visualization
  addLayersControl(
    overlayGroups = c("Confirmations","Deaths"),
    options = layersControlOptions(collapsed = FALSE)
  )

```

## Objective 2 

This objective had to do with creating a chart with kable to show the order of countries in descending order by deaths/confirmations. The first thing I did was make new data frames with only the information I needed. I then called **knitr::kables** and created two kable charts in the list. I added captions and column names to create the two charts next to each other. 
```{r Obj2}
#sets a new data frame for the confirmation data set 
newerConfirm <- select(newConfirm, Country.Region, totalConfirms)
newerConfirm <- data.frame(newerConfirm %>% arrange(desc(totalConfirms)))

#sets a new data frame for the death data set
newerDeath <- select(newDeath, Country.Region, totalDeaths)
newerDeath <- data.frame(newerDeath %>% arrange(desc(totalDeaths)))



#Sets two charts and sets them next to each other
knitr::kables(
  list(
    #First chart
    knitr::kable(
      newerConfirm, col.names = c("Country","Count"), valign='t', caption = "Confirms"
    ),
    #Second chart
    knitr::kable(newerDeath, digits = 0, valign='t', col.names = c("Country","Count"), caption = "Death")
    
    
  ),
  
 
)


```

## Objective 3

After reading the top tables, you are stunned! The US overtakes every other country in terms of COVID-19 confirmations. As such, you are concerned about the state you live in and would like to understand how COVID-19 events have shaped the trajectory of the disease. Create two scatter plots to gain a better understanding. The first scatter plot should be California’s trajectory for confirmations. The second scatter plot should show California’s top three city trajectories for confirmations. You are interested in studying how the vaccine affected the number of confirmations. The Moderna vaccine was first available as an emergency use authorized (EUA) vaccine and required two shots spaced six weeks apart. Indicate on the plots the day the second dosage was given to those that received the first dosage the day Moderna was EUA (January 29th, 2021). As a diligent scientist that knows that new COVID variants have mutations in the spike protein (the region that the vaccine was developed for), you also want to study how confirmation rates change as new variants become the dominant infectious strain. Indicate on the plots when the delta and omicron variants became the dominant strain in California (May 11th, 2021 and November 26th, 2021 respectively). In the example below, the function plot_grid from the R package cowplot was to organize the graphs into a grid to more easily compare statewide vs top city plots.

```{r}
# Filter data for California
california_data <- confirmedGlobal %>%
  filter(Country.Region == "US" & Province.State == "California")

# Filter data for top three cities in California
top_cities_data <- california_data %>%
  arrange(desc(Latest.Confirmed)) %>%
  head(3)

# Convert Date column to Date type
california_data$Date <- as.Date(california_data$Date)

# Create a scatter plot for California's trajectory
plot_california <- ggplot(california_data, aes(x = Date, y = Latest.Confirmed)) +
  geom_point() +
  geom_vline(xintercept = as.Date("2021-01-29"), linetype = "dashed", color = "blue", size = 1) + # EUA of Moderna vaccine
  geom_vline(xintercept = as.Date("2021-05-11"), linetype = "dashed", color = "green", size = 1) + # Dominance of Delta variant
  geom_vline(xintercept = as.Date("2021-11-26"), linetype = "dashed", color = "red", size = 1) + # Dominance of Omicron variant
  labs(title = "California's COVID-19 Confirmed Cases Trajectory",
       x = "Date",
       y = "Confirmed Cases") +
  theme_minimal()

# Create scatter plots for top three city trajectories
plot_top_cities <- lapply(1:3, function(i) {
  ggplot(top_cities_data[i, ], aes(x = Date, y = Latest.Confirmed)) +
    geom_point() +
    geom_vline(xintercept = as.Date("2021-01-29"), linetype = "dashed", color = "blue") + # EUA of Moderna vaccine
    geom_vline(xintercept = as.Date("2021-05-11"), linetype = "dashed", color = "green") + # Dominance of Delta variant
    geom_vline(xintercept = as.Date("2021-11-26"), linetype = "dashed", color = "red") + # Dominance of Omicron variant
    labs(title = paste("Top City", i, "COVID-19 Confirmed Cases Trajectory"),
         x = "Date",
         y = "Confirmed Cases") +
    theme_minimal()
})

# Organize plots into a grid
grid_plot <- cowplot::plot_grid(plot_california, plot_top_cities[[1]], plot_top_cities[[2]], plot_top_cities[[3]], ncol = 2)

# Display the grid plot
grid_plot

```


## Objective 4

Although these plots do not tell the whole story, they are great for helping us determine where to look. Different cities may have different populations, population densities, cultural discrepancies, compliance, and city regulations to name a few. We will explore the role of population on these metrics using visualizations. Arrange two scatter plots using cowplot’s plot_grid to show the relationship between population and confirmed counts as well as death counts and confirmed counts. You will need to use a log transform on all variables to show such a relationship. Please consult the example below for an idea of what this may look like. From these graphs we can see that population greatly affects confirmations and deaths. This coincides with our plots above as Los Angeles’s population is 301% greater than San Diego’s population and 406% greater than Riverside’s population!

```{r}
# Filter data for California
california_confirmed_data <- confirmedGlobal %>%
  filter(Country.Region == "US" & Province.State == "California")

# Filter data for population of California
california_population <- california_confirmed_data %>%
  select(Province.State, Population) %>%
  distinct()

# Merge population data with confirmed cases and deaths data
california_confirmed_data <- merge(california_confirmed_data, california_population, by = "Province.State")

# Log transform the variables
california_confirmed_data$log_Confirmed <- log(california_confirmed_data$Latest.Confirmed)
california_confirmed_data$log_Deaths <- log(california_confirmed_data$Latest.Deaths)
california_confirmed_data$log_Population <- log(california_confirmed_data$Population)

# Create scatter plot for relationship between population and confirmed counts
plot_population_confirmed <- ggplot(california_confirmed_data, aes(x = log_Population, y = log_Confirmed)) +
  geom_point() +
  labs(title = "Relationship Between Population and Confirmed Counts",
       x = "Log Population",
       y = "Log Confirmed Counts") +
  theme_minimal()

# Create scatter plot for relationship between death counts and confirmed counts
plot_deaths_confirmed <- ggplot(california_confirmed_data, aes(x = log_Confirmed, y = log_Deaths)) +
  geom_point() +
  labs(title = "Relationship Between Confirmed Counts and Death Counts",
       x = "Log Confirmed Counts",
       y = "Log Death Counts") +
  theme_minimal()

# Organize plots into a grid
grid_plot <- plot_grid(plot_population_confirmed, plot_deaths_confirmed, ncol = 2)

# Display the grid plot
grid_plot

```
